---
title: "Chapter 6: Reading Data into R"
output: html_notebook
---

6.1 Reading CSVs
```{r}

theURL <- "http://www.jaredlander.com/data/Tomato%20First.csv"
tomato <- read.table(file=theURL, header=TRUE, sep=',')
head(tomato)

x <- 10:1
y <- -4:5
q <- c("Hockey", "Football", "Baseball", "Curling", "Rugby", "Lacrosse", "Basketball", "Tennis", "Cricket", "Soccer")
theDf <- data.frame(First=x, Second=y, Sport=q, stringsAsFactors=FALSE)
theDf
theDf$Sport

```
6.2 Excel Data
Convert .xslx data into .csv file first, and then use header.table. That is the fastest way. There is also a RODBC package if access to a database via DSN connection is available.

6.3 Reading from Databases
```{r}

# warning: takes too long and will crash R session
install.packages("RODBC")
require('RODBC')
db <- odbcConnect("QV Training")
ordersTable <- sqlQuery(db, "SELECT * FROM Orders", stringsAsFactors=FALSE)
detailsTable <- sqlQuery(db, "SELECT * FROM [Order Details]", stringsAsFactors=FALSE)
longQuery <- "SELECT * FROM Orders, [Order Details] WHERE Orders.OrderID = [Order Details].OrderID"
detailsJoin <- sqlQuery(db, longQuery, stringsAsFactors=FALSE)

```

6.4 Data from Other Statistical Tools
```{r}

# 'foreign' package contains tools and funcitons similar to those in read.table
install.packages('foreign')
require('foreign')

# more functions
read.spss #SASS
read.dta #Stata
read.ssd #SAS
read.octave # Octave
read.mtp # Minitab
read.systat # Systat

```

6.5 R Binary Data
```{r}

fp <- '/Users/fennec2000/OneDrive - University of Florida/Computer Science/Projects/R Projects/R-Crash-Course/src/tomato.rdata'
save(tomato, file=fp)
rm(tomato)
head(tomato)
load(file=fp, verbose=TRUE)
head(tomato)

```

```{r}

n <- 20
r <- 1:10
w <- data.frame(n, r)
n
r
w

fp <- '/Users/fennec2000/OneDrive - University of Florida/Computer Science/Projects/R Projects/R-Crash-Course/src/multiple.rdata'
save(w, file=fp)
rm(n, r, w)
load(file=fp, verbose=TRUE)
w

```

6.6 Data Included with R
```{r}

# some data sets come as default in certain R packages
require(ggplot2)
data(x=diamonds)
head(x=diamonds)

```

6.7 Extract Data from Websites
6.7.1 Simple HTML Tables
```{r}

install.packages("XML")
require(XML)
theURL <- "http://www.jaredlander.com/2012/02/another-kind-of-super-bowl-pool/"
bowlPool <- readHTMLTable(theURL, which=1, header=FALSE, stringsAsFactors=FALSE)
bowlPool

```

6.7.2 Scraping Web Data
*see section 13.14






